\section{Status}
\label{sec:status}

\subsection{Implemented}

The following capabilities are implemented and functional as of the document date:

\begin{itemize}
  \item SQLite schema with \texttt{roots}, \texttt{files}, \texttt{nodes},
        \texttt{edges}, \texttt{refs}, \texttt{nodes\_fts},
        \texttt{content\_fts}, and \texttt{content\_fts\_tracker}.
  \item Tree-sitter parsing for 13 languages: C, C++, C\#, TypeScript,
        JavaScript, Python, Rust, Java, Go, Bash, SQL, YAML, Lua.
  \item Symbol extraction (definitions) and call-edge extraction for all
        supported languages.
  \item All 27 MCP tools registered and functional (24 core + 3 workspace).
  \item Incremental indexing via mtime + xxHash fingerprinting.
  \item Filesystem watcher (FSEvents / inotify / ReadDirectoryChangesW)
        for live re-indexing on file changes.
  \item Multi-root workspace with ID offset scheme and cascade removal.
  \item Freshness policy: \texttt{eager}, \texttt{normal}, \texttt{lazy},
        \texttt{off}.
  \item \texttt{QueryCache}: prepared-statement cache cleared on reindex.
  \item Staleness detection via \texttt{.git/HEAD} mtime tracking.
  \item Bulk mode for repositories $> 1\,000$ files (drop/rebuild indexes).
  \item Auto-editor configuration (\texttt{codetopo init}) for VS Code,
        Cursor, Windsurf, and Claude Desktop.
\end{itemize}

\subsection{Known limitations}

\begin{itemize}
  \item \textbf{Approximate call edges.} Cross-file call resolution is
        AST-derived without type resolution. Overloaded functions, virtual
        dispatch, and generic instantiations are not distinguished. Tools
        are named \texttt{callers\_approx} / \texttt{callees\_approx} to
        make this explicit.
  \item \textbf{No taint or dataflow analysis.} Security and refactoring
        workflows that require tracing data flow between symbols are not
        supported.
  \item \textbf{No formal LSP endpoint.} codetopo speaks MCP. Editors that
        rely on the Language Server Protocol~\cite{microsoft2016lsp} cannot
        use codetopo directly as a language server.
  \item \textbf{No remote or team-shared index.} The index is per-developer
        local. There is no shared index server or sync protocol.
  \item \textbf{No symbol-importance ranking.} FTS5 BM25 provides text
        relevance; there is no PageRank-style scoring over the call graph.
\end{itemize}

\subsection{Future work}

\begin{description}
  \item[Type-resolved call edges.] Optionally invoke a language server for
        type information to improve call-edge accuracy for overloaded and
        generic code.
  \item[Additional languages.] Add extractors for Ruby, Kotlin, Swift, PHP,
        Zig, and Haskell. Grammars exist for all of these in the
        tree-sitter~\cite{brunsfeld2018treesitter} ecosystem.
  \item[LSP adapter.] Expose an LSP endpoint backed by the codetopo index
        to expand editor coverage beyond the VS Code / Cursor / Windsurf family.
  \item[PageRank on the call graph.] Weight symbol search results by structural
        importance (in-degree on the call graph) in addition to BM25 text
        relevance.
  \item[Remote/shared index.] Implement a sync protocol so that large teams can
        share a pre-built index rather than indexing individually.
\end{description}
